Argennon: A Scalable Smart Contract Platform

### 2022-03-30 tex/sections/avm.tex
Update Chapter 1
This is the first draft.

Signed-off-by: aybehrouz <behrouz_ayati@yahoo.com>

An Argennon Standard Representation (ASR) is a programming language for describing argennon applications, optimized
for the architecture and requirements of the Argennon platform.
Argennon supports two standard representations: one is a high level text based language which needs
costly compilation before being executed on a hardware machine. The other one is a low level binary representation which
usually can be executed, with minimal pre-processing, by a JIT compiler or an emulator. The high level
language is intended for preserving high level information of applications logic to facilitate
platform specific compiler optimization at host nodes. On the other hand, the low level language enables efficient
execution of applications that are not frequently used.

The state of an Argennon application is stored in byte addressable finite arrays of memory called
\emph{heap chunks}. An application can have several heap chunks with different sizes, and can remove or
resize its heap chunks or allocate new chunks. Every chunk belongs to exactly one application and can only be modified
by its owner. In addition to heap chunk, every application also has an amount of non-persistent local memory for
storing temporary data.

The Argennon Smart Contract Execution Environment can be seen as a machine for executing http requests, producing
their http response and updating related heap
chunks. The AscEE executes requests sequentially \footnote{Actually requests are executed in parallel but by
performing data dependency analysis the result is guaranteed to be identical with sequential execution of
requests.} and each request has a separate \emph{execution session}. Execution sessions are separate sessions of
executing smart contract's code by the Argennon Smart Contract Execution Environment. The state of an execution
session will be
destroyed at the end of the session and only the state of heap chunks is preserved. If a session fails and does not
complete normally, it will not have any effect on the heap chunks.

During an execution session an application can make http requests to other applications. Those requests will not start
a new execution session and will be executed within the current session. In AscEE making a http request to an
application is similar to a function invocation therefore we also refer to it as an application call.

@@

Argennon has four primitive prefix trees:
\emph{applications, accounts, local} and \emph{varUint}.
All these trees are in base 256, with the maximum height
of 8. An Argennon identifier may be simple or compound. A simple identifier is generated using a single trie, while a
compound identifier is generated by concatenating prefix codes generated by two or more tries:

@@

In Argennon the shorter prefix codes are assigned to more active accounts and smart contracts which tend to own more
data objects in the system. The prefix trees are designed by analyzing empirical data to make sure the number
of leaves in each level is chosen appropriately.

@@

When an application allocates a new heap chunk, the identifier of the new chunk is not generated by
the AscEE. Instead, the application can choose an identifier itself \footnote{provided it has a correct format}. This
is an important feature of the AscEE's heap, which allows applications to use the AscEE's heap as a dictionary (map)
data structure.
Since the \texttt{chunkID} is a prefix code, any application has its own identifier space, and an application
can easily generate unique identifiers for its chunks.

A heap chunk can be considered as a continuous
array of bytes. Every chunk has a size: \texttt{chunkSize} and a size upper bound: \texttt{sizeUpperBound}. The value of
\texttt{chunkSize} can be determined uniquely at
the start of
every execution session, and it may be updated during the session like a normal memory location. On the other hand,
the value of \texttt{sizeUpperBound} is constant for every block of the blockchain and is proposed by the block
proposer.

@@

Memory locations inside a chunk can only be accessed through access blocks. An access block is defined on a chunk
and has an offset and a size and determines accessible memory locations inside a chunk. Multiple access blocks can
be defined on a chunk, but they must be non overlapping. Access blocks are byte addressable and can have different
access types:

\begin{itemize}
    \item \texttt{read\_only}: only allows read and check operations.
    \item \texttt{writable}: memory locations inside this access block can not be modified.
    \item \texttt{check\_only}: only allows check operations. These operations query the persistence
    status of memory locations.
    \item \texttt{additive}: only allows addition operations without overflow checking. Note that the content of these
    access blocks cannot be read.
\end{itemize}

@@

The value of chunkSize can be modified during an execution session. However, the new size values can only be
increasing or decreasing. More precisely, if a request declares that it wants to expand (shrink) a chunk, it can only
increase (decrease) the value of chunkSize and any specified value for chunkSize during the execution session, needs
to be greater (smaller) than the previous value of the chunk size. Any request that wants to expand (shrink) a chunk
needs to specify a max size (min size). The value of chunkSize can not be set higher (lower) than this value.

@@

There is no way for an application to query the chunk capacity. As a result, in the view of an application, accessing
offsets higher than chunkSize results in undefined behaviour, while the behaviour is well-defined for validators.
This enables validators to determine the validity of an offset at the start of the block validation in a parallelized
preprocessing phase without executing requests.

@@

@@

Completing an execution session needs computational resources. The amount of resources used by an execution session
should be monitored, otherwise a malicious user would be able to easily spam and exhaust resources of the execution
environment. Resource usage can be measured per session or per application call.

The AscEE has two type of execution sessions: \emph{optimistic} and \emph{monitored}. Resource usage of an optimistic
session is always measured per session and default pre-defined resource caps are used. On the other hand in a monitored
session some resources are measured per application and some
caps are determined dynamically by the transaction. An optimistic session must always
complete normally, a monitored session is allowed to either complete normally or abruptly.

Obviously resource monitoring is easier and more efficient for optimistic sessions. However, if
an optimistic session violates its resource caps, determining the point of failure requires precise resource
measurement. Note that for implementing optional decoupling principle, we need to determine the exact application
which has failed in the call chain. For example assume that an optimistic session containing an application call
violates a 2 milliseconds execution time cap. If in the caller application the call happens exactly after 2
milliseconds of execution, a small fluctuation in the execution time measurement can
change the point of failure between the called and the main application. This can introduce nondeterministic
behaviour which makes block validation impossible. On the other hand, for a monitored session this is not an issue
because resource measurements and caps are defined per application, and we make sure that resource caps are not too
small and can be measured easily.

Different computational resources are measured and monitored during an AscEE session:
\begin{itemize}
    \item \textbf{execution time}:
    the amount of cpu time that is required for executing a session or an application. The execution time of a session
    is measured in \emph{AscEE clocks}. One AscEE clock is defined as 1/1000 of the amount
    of cpu time needed for executing a predefined standard application which is used for benchmarking a host's
    performance.

@@

    Monitored sessions perform per application cpu-time measurement, and every application call during a monitored
    session has a separate \texttt{maxClocks} value. This value determines the maximum amount of time that the cpu
    can be used for executing that particular application call. It should be noted that this cpu timer is paused when an
    application makes a call to another application, until the control returns to the caller. An application call
    needs at least 100 clocks and if the value of its \texttt{maxClock} is lower than this value the call will be
    considered a failed call.

    Each application call has some amount of \texttt{externalClocks}. When an application makes a request to another
    application it has to \emph{forward} a portion of its external clocks to the called application. This amount
    will determine the value of maxClocks for the called application.

    The amount of external clocks of an application
    call is defined to be \(2/3\) of its \texttt{maxClocks}. As a result, the total number of clocks of a monitored
    session is always less than \(3 \times \texttt{maxClocks}\) of the root application. The value of
    \texttt{maxClocks} for
    the root application call is determined by the transaction.

@@

    Optimistic sessions measure local memory usage per session and enforce a protocol-defined cap on the total amount of
    used local memory in the session. Monitored sessions measure local memory usage per
    application call and enforce a protocol defined-cap per application call. An application call which tries to use
    more local memory than the cap will fail.
    \item \textbf{heap access list}:
    every session can only access heap locations that are declared in its access list. In addition,
    resizing heap chunks can only be done in the range of the pre-declared lower bound and upper bound.
    \item \textbf{applications list}:
    a session may only make requests to applications that are declared in its application list.
    \item \textbf{call depth}:
    during a session the number of nested application calls can not be more than a threshold. This threshold is
    determined by the Argennon protocol.
    \item active differed calls:
    \item virtual signatures:
    \item number of locks

\end{itemize}

@@

Attachments of a request is a list of request identifiers of the current block that are "attached" to the request.
That means, for validating that request a validator first needs to "inject" the digest of attached requests into the
httpRequest field of the request.

\section{Arithmetics}\label{sec:arithmetics}

The Argennon Virtual Machine supports signed integer and signed floating point operations. The Argennon Virtual
Machine does not support any type of unsigned arithmetics. All arithmetic operations in the Argennon Virtual Machine
are checked and any type of overflow or underflow will cause a catchable exception to be thrown.

The Argennon Virtual Machine provides low level reentrancy protection by defining a dedicated
instruction: \texttt{enter}. When a smart contract executes the \texttt{enter} instruction, it
can not execute that instruction again as long
as the method executing the instruction completes (abruptly or normally). If a smart contract tries to execute
the \texttt{enter} instruction while the method executing that instruction for the first time has not yet completed,
an exception will be thrown.

Smart contracts can completely prevent reentrancy by executing the \texttt{enter} instruction as the first instruction
of their \texttt{dispatcher} method.

In blockchain applications, we usually need to authorize certain operations. For example, for sending an asset
from a user to another user, first we need to make sure that the sender has authorized this operation. The
Argennon Virtual Machine has no built-in mechanism for authorizing operations, but it provides a rich set of
cryptographic instructions for validating signatures and cryptographic entities. By using these instructions and
passing cryptographic signatures as parameters to methods, a programmer, having users' public keys, can implement
the required logic for authorizing any operation.

\note{Authorization by explicit signatures eliminates the need for approval methods or call back patterns.}

In addition to cryptographic signatures, the Argennon Virtual Machine provides instructions that enable
smart contracts to issue \emph{virtual signatures}. A smart contract can virtually sign a message for a target
smart contract by \texttt{v\_sign} instruction. When a smart contract
virtually signs a message for a target, that message is stored in the internal memory of the AVM, associated with
its signer and the target. The application that is the target of the signature can verify the message
by \texttt{v\_verify} instruction. After verifying the signature, it will be removed from the AVM's internal memory
to make sure a signature will not be used multiple times. The virtual signatures stored in the AVM's internal memory
last for one execution session.

\note{The Argennon virtual machine has no instructions for issuing cryptographic signatures.}

\section{The Argennon Standard Library}\label{sec:asl}

As we explained in Section~\ref{subsec:method-invocation} by using \texttt{invoke\_internal} instruction, a smart
contract can invoke methods of another smart contract in its own context. AVM smart contracts use this instruction to
invoke methods of the AVM standard library. Methods of the AVM standard library are stored
with the \texttt{applicationID} zero in the AVM method area. This implies that the AVM standard library is actually
a part of the root smart contract.

In Argennon, the root smart contract is an updatable smart contract, which can be updated by the Argennon governance
system. This means that bugs or security vulnerabilities in the AVM standard library could be quickly patched and
smart contracts could benefit from bugfixes and improvements of the AVM standard library even if they are
non-updatable. Many important and useful functionalities,
such as fungible and non-fungible assets, access control mechanisms,
and general purpose DAOs are implemented in the Argennon standard library.

@@

>>>>

An Argennon Standard Representation (ASR) is a programming language for describing argennon applications, optimized
for the architecture and requirements of the Argennon platform.
Argennon supports two standard representations: one is a high level text based language which needs
costly compilation before being executed on a hardware machine. The other is a low level binary representation which
usually can be executed, with minimal pre-processing, by a JIT compiler or an emulator. The high level
language is intended for preserving high level information of applications logic to facilitate
platform specific compiler optimization at host nodes. On the other hand, the low level language is designed for
efficient execution of applications that are not frequently used.

The state of an Argennon application is stored in byte addressable finite arrays of memory called
\emph{heap chunks}. An application can have several heap chunks with different sizes, and can remove or
resize its heap chunks or allocate new chunks. Every chunk belongs to exactly one application and can only be modified
by its owner. In addition to heap chunks, every application also has an amount of non-persistent local memory for
storing temporary data.

\section{Execution Sessions}\label{sec:sessions}

The Argennon Smart Contract Execution Environment can be seen as a machine for executing http requests, producing
their http response and updating related heap chunks. The AscEE executes requests sequentially \footnote{Actually
requests are executed in parallel but by performing data dependency analysis the result is guaranteed to be identical
with sequential execution of requests.} and each request has a separate \emph{execution session}. Execution sessions
are separate sessions of executing smart contract's code in order to fulfill \emph{external} http requests. An
external request is a request that is not made by another Argennon application.

The state and data of an execution session will be
destroyed at the end of the session and only the state of heap chunks is preserved. If a session fails and does not
complete normally, it will not have any effect on any heap chunks.

During an execution session an application can make \emph{internal} http requests to other applications. Those
requests will not start a new execution session and will be executed within the current session. In AscEE making a
http request to an application is similar to a function invocation. For that reason, we also refer to internal requests
as application calls.

@@

Argennon has four primitive prefix trees:
\emph{applications, accounts, local} and \emph{varUint}.
All these trees are in base 256, with the maximum height
of 8.

An Argennon identifier may be simple or compound. A simple identifier is generated using a single trie, while a
compound identifier is generated by concatenating prefix codes generated by two or more tries:

@@

In Argennon the shorter prefix codes are assigned to more active accounts and applications which tend to own more
data objects in the system. The prefix trees are designed by analyzing empirical data to make sure the number
of leaves in each level is chosen appropriately.

@@

When an application allocates a new heap chunk, the identifier of the new chunk is not generated by
the AscEE. Instead, the application can choose an identifier itself \footnote{provided it has a correct format}. This
is an important feature of the AscEE's heap, which allows applications to use the AscEE's heap as a dictionary (map)
data structure.
Since the \texttt{chunkID} is a prefix code, any application has its own identifier space, and an application
can easily find unique identifiers for its chunks.

A heap chunk can be considered as a continuous
array of bytes. Every chunk has a size: \texttt{chunkSize} and a size upper bound: \texttt{sizeUpperBound}. The value of
\texttt{chunkSize} can be determined uniquely at
the start of
every execution session, and it may be updated during the session like a normal memory location. On the other hand,
the value of \texttt{sizeUpperBound} is constant for every block of the blockchain and is proposed by the block
proposer. \texttt{sizeUpperBound} indicates the upper bound of the value of \texttt{chunkSize} for a chunk in a block.
A validator can use this value to safely allocate memory for a chunk.

@@

Memory locations inside a chunk can only be accessed through access blocks. An access block is defined on a chunk
and has an offset and a size and determines accessible memory locations inside a chunk. Multiple access blocks can
be defined on a single chunk, but they must be non overlapping. Access blocks are byte addressable and can have
different
access types:

\begin{itemize}
    \item \texttt{read\_only}: only allows read and check operations.
    \item \texttt{writable}: allows reading and writing.
    \item \texttt{check\_only}: only allows check operations. These operations query the persistence
    status of a memory location.
    \item \texttt{additive}: only allows addition operations without overflow checking. Note that the content of these
    access blocks cannot be read.
\end{itemize}

@@

The value of \texttt{chunkSize} can be modified during an execution session. However, the new size values can only be
increasing or decreasing. More precisely, if a request declares that it wants to expand (shrink) a chunk, it can only
increase (decrease) the value of \texttt{chunkSize} and any specified value during the execution
session, needs to be greater (smaller) than the previous value of the chunk's size. Any request that wants to expand
(shrink) a chunk needs to specify a max size (min size). The value of \texttt{chunkSize} can not be set higher
(lower) than this value.

@@

There is no way for an application to query \texttt{sizeUpperBound} of a chunk. As a result, for an application,
accessing offsets higher than \texttt{chunkSize} results in undefined behaviour, while the behaviour is well-defined
in the view of validators.
This enables validators to determine the validity of an offset at the start of the block validation in a parallelized
preprocessing phase without actually executing requests.

@@

Attachment of a request is a list of request identifiers of the current block that are "attached" to the request.
That means, for validating that request a validator first needs to "inject" the digest of attached requests into the
http request.

In blockchain applications, we usually need to authorize certain operations. For example, for sending an asset
from a user to another user, first we need to make sure that the sender has authorized this operation.

The AscEE uses \emph{authenticated message passing} for authorizing operations. In this method every execution
session has a set of authenticated messages. Applications explicitly pass those messages for authorizing operations.
Authentication of messages can be done using any type of cryptographic signature scheme in the preprocessing phase of
request execution. For example, we could use BLS aggregate signature to authenticate all messages of a block in bulk.

Moreover, application use built-in functions of AscEE to generate authenticated messages in run-time. This enables an
application to authorize an operation even for an application that is not directly called.

\note{Authorization by explicit messages and signatures eliminates the need for approval methods or call back patterns.}

In addition to authenticated messages the AscEE provides a set of
cryptographic functions for validating signatures and cryptographic entities. By using these functionalities and
passing cryptographic signatures as parameters to methods, a programmer, having users' public keys, can implement
the required logic for authorizing operations.

\note{The Argennon Execution Environment has no instructions for issuing cryptographic signatures.}

The Argennon Smart Contract Execution Environment provides optional low level reentrancy protection by providing low
level \emph{entrance locks}. When an application acquires an entrance lock it cannot acquire that lock again and trying
to do so will result in a revert. The entrance lock of an application will be released when the application explicitly
releases its entrance lock or when the call that had acquired the lock completes.

The AscEE reentrancy protection mechanism is optional. An application can allow reentrancy, it can protect only certain
areas of its code, or can completely disallow reentrancy.

\ldots

@@

Completing an execution session requires computational resources. The amount of resources used by an execution session
should be monitored, otherwise a malicious user would be able to easily spam and exhaust resources of the execution
environment. Resource usage can be measured per session or per application call.

The AscEE has two type of execution sessions: \emph{optimistic} and \emph{monitored}. Resource usage of an optimistic
session is always measured per session and default pre-defined resource caps are used. On the other hand in a monitored
session, some resources are measured per application call and some
caps are determined dynamically by the external http request (i.e.\ transaction). An optimistic session must always
complete normally, a monitored session is allowed to either complete normally or abruptly.

Obviously resource monitoring is easier and more efficient for optimistic sessions. However, if
an optimistic session violates its resource caps, determining the point of failure requires precise resource
measurement. Note that for implementing optional decoupling principle, we need to determine the exact application
which has failed in a call chain. For example assume that an optimistic session containing an application call
violates a 2 milliseconds execution time cap. If in the caller application the call happens exactly after 2
milliseconds of execution, a small fluctuation in the execution time measurement can
change the point of failure between the called and the main application. This can introduce nondeterministic
behaviour which could make block validation impossible. On the other hand, for a monitored session this is not an issue
because resource measurements and caps are defined per application, and we can make sure that resource caps are not too
small and can be measured easily.

Different computational resources are measured and monitored during an AscEE session:
\begin{itemize}
    \item \textbf{execution time}:
    the amount of cpu time that is required for executing a session or an application call. The execution time of a
    session is measured in \emph{AscEE clocks}. One AscEE clock is defined as 1/1000 of the amount
    of cpu time needed for executing a predefined standard application which is used for benchmarking a host's
    performance.

@@

    Monitored sessions perform per application call cpu-time measurement, and every application call during a monitored
    session has a separate \texttt{maxClocks} value. This value determines the maximum amount of time that the cpu
    can be used for executing that particular application call. It should be noted that the cpu timer is paused when
    the application makes a call to another application, and is resumed when the control returns. An application call
    needs at least 100 clocks and if the value of its \texttt{maxClock} is lower than this value the call will be
    considered a failed call.

    Each application call has some amount of \texttt{externalClocks}. When an application makes a request to another
    application it has to \emph{forward} a portion of its external clocks to the called application. This amount
    will determine the value of \texttt{maxClocks} for the called application.

    The amount of external clocks of an application
    call is defined to be \(2/3\) of its \texttt{maxClocks}. As a result, the total number of clocks of a monitored
    session is always less than \(3 \times \texttt{maxClocks}\) of the root application call. The value of
    \texttt{maxClocks} for
    the root application call is determined by the external request (transaction).

@@

    Optimistic sessions measure local memory usage per session and enforce a protocol-defined cap on the total amount of
    used local memory in the session. Monitored sessions measure local memory usage per
    application call and enforce a protocol defined-cap per application call. An application call which tries to use
    more local memory than the cap will fail.
    \item \textbf{heap access list}:
    every session can only access heap locations that are declared in its access list. In addition,
    resizing heap chunks can only be done in the range of the pre-declared lower bound and upper bound.
    \item \textbf{applications list}:
    a session may only make requests to applications that are declared in its application list.
    \item \textbf{call depth}:
    during a session the number of nested application calls can not be more than a threshold. This threshold is
    determined by the Argennon protocol.
    \item \textbf{active differed calls}:
    \item \textbf{virtual signatures}:
    \item \textbf{number of entrance locks}:
\end{itemize}

@@

\section{The ArgC Language}\label{sec:the-argc-language}

\subsection{The ArgC Standard Library}\label{sec:asl}

An application can invoke methods of the ArgC Standard Library (ASL) in its own context. Methods of the ArgC standard
library are stored as a special application in the Argennon blockchain.
a part of the root smart contract.
In Argennon, some applications (smart contracts) are updatable. The ArgC Standard Library is an updatable smart
contract which can be updated by the Argennon governance
system. This means that bugs or security vulnerabilities in the ArgC Standard Library could be quickly patched and
applications could benefit from bugfixes and improvements of the ArgC Standard Library even if they are
non-updatable. Many important and useful functionalities,
such as fungible and non-fungible assets, access control mechanisms,
and general purpose DAOs are implemented in the ArgC Standard Library.

@@

\section{Parallel Request Execution}\label{sec:concurrency}
\input{sections/concurrency}
